Puck Predictions: Unraveling the NHL Game Forecasting Riddle

### 2024-04-12 Latex/draft.tex
Pushing final read through.

Our project seeks a similar outcome to the research mentioned above: predict the outcomes of NHL games. Not only this,
but we seek to provide live, acturate, win probabilities for any given game state. Despite the simplicity of the problem statement, 
the solution is not so straightforward. The NHL provides fast-paced games with many events
occuring in quick succession. Our goal is to use this abundance of data and new approaches to build upon previous research.

@@

The models were each trained on prior seasons, and tested on the most recent 2023-2024 season. This split was chosen because in practice our model would have access to historical data but not have access to the current season's data as it happens.

@@

To simulate hockey games, we created a Markov Chain where the states are a tuple of three consecutive
events that occurred. For example, if the home team won a faceoff, lost the puck, and 
then the away team shot the puck and missed, the markov chain would look like Table~\ref{tab:markov_chain_sample}.

@@

To simulate a game, we performed this Monte Carlo Algorithm (see Algorithm~\ref{alg:monte_carlo_algorithm} in Appendix~\ref{mcmc_app})
that acts as a random walk through the Markov Chain (Note that the {\tt KDE\_times()} is a KDE model fit on the amount of seconds that transpired between 
each hockey event for all NHL games over the course of 13 years).
% \begin{algorithm}
%     \caption{Simulation Algorithm}
%     \begin{algorithmic}[1]
%         \State $\text{time\_remaining} \gets 3600$
%         \Comment{NHL games are 3600 seconds}
%         \State $s_0 \gets "\#"$
%         \Comment{The "\#" symbol represents the start of a game}
%         \State $s_1 \gets "\#"$
%         \State $s_2 \gets "\#"$
%         \State $\text{event\_counts} \gets \text{dictionary()}$

@@

Overall, the Bayesian network performed well, and was able to produce realistic win probabilities. Because it defined a joint distribution using a DAG and
then used that distribution to fit the data, it was able to correctly predict accuracies using the few features provided. However, the Bayesian network struggled to capture the intricate dependencies
of factors other than goals that could affect the win probabilities, so the predicted probabilities are little more than an over all probability calculation of goals and time reamining given historical data.
We see this in the probability graph in Figure~\ref{fig:bayesian_graph}, where the changes in probability correspond to the goals scored in the game.

@@

To evaluate our simulation's effectiveness at accurately modeling actual hockey 
games, we combined a dataset with the final event counts for 1500 actual hockey 
games and for 1500 simulated games. We then performed two experiments. First, We 
trained a KMeans, Logistic Regression, RandomForest and XGBoost model on this 
dataset with default parameters, to see 
if they could classify games as simulated or real. With a 70-30 train-test split, 
the accuracies on the test set were 47.9\%, 48.5\%, 67.5\% and 69.3\% respectively. 
Second, we fit and transformed this dataset using PCA, t-SNE and UMAP 
dimensionality reductions using various perplexity and neighbor hyperparameters to see 
if these algorithms clustered the synthetic games and actual games in 
different clusters. As shown below in Figure~\ref{fig:simulation_v_actual}, the simulated
games and actual games are all clustered together. From these two experiments, we conclude
that our simulator effectively models live NHL games.

@@

\subsection{Method Comparison}
Each of our methods had one general purpose: predict the winner of the game regardless of the point of the game. 
To evaluate our models, we randomly selected 100 games outside of the training data. 
To ensure that the test set represented the actual data, we selected the test set with 50 home wins and 50 away wins, 
and we also made sure there was an even representation of each of the 32 NHL teams. 
We then used each model to predict a winner at various points throughout each game. 
From these predictions on the 100 games, we computed each model's accuracy for each point in the game.

>>>>

Our project seeks a similar outcome to the research mentioned above: predict the outcomes of NHL games. Not only this,
but we seek to provide live, accurate, win probabilities for any given game state. Despite the simplicity of the problem statement, 
the solution is not so straightforward. The NHL provides fast-paced games with many events
occuring in quick succession. Our goal is to use this abundance of data and new approaches to build upon previous research.

@@

The models were each trained on the 2021-2023 seasons (totaling two seasons), and tested on the most recent 2023-2024 season. This split was chosen because in practice our model would have access to historical data but not have access to the current season's data as it happens.

@@

To simulate hockey games, we created a Markov Chain where the states are a tuple of three consecutive
events that occurred. For example, if the home team won a faceoff, lost the puck, and 
then the away team shot the puck and missed, the next potential event in the Markov chain would look like Table~\ref{tab:markov_chain_sample}.

@@

To simulate a game, we performed a Monte Carlo algorithm (see Algorithm~\ref{alg:monte_carlo_algorithm} in Appendix~\ref{mcmc_app})
that acts as a random walk through the Markov chain (note that the {\tt KDE\_times()} is a KDE model fit on the amount of seconds that transpired between 
each hockey event for all NHL games over the course of 13 years).
% \begin{algorithm}
%     \caption{Simulation Algorithm}
%     \begin{algorithmic}[1]
%         \State $\text{time\_remaining} \gets 3600$
%         \Comment{NHL games are 3600 seconds}
%         \State $s_0 \gets "\#"$
%         \Comment{The "\#" symbol represents the start of a game}
%         \State $s_1 \gets "\#"$
%         \State $s_2 \gets "\#"$
%         \State $\text{event\_counts} \gets \text{dictionary()}$

@@

Overall, the Bayesian network performed well and was able to produce realistic win probabilities. Because it defined a joint distribution using a DAG and
then used that distribution to fit the data, it was able to correctly predict accuracies using the few features provided. However, the Bayesian network struggled to capture the intricate dependencies
of factors other than goals that could affect the win probabilities, so the predicted probabilities are little more than an over all probability calculation of goals and time reamining given historical data.
We see this in the probability graph in Figure~\ref{fig:bayesian_graph}, where the changes in probability correspond to the goals scored in the game.

@@

To evaluate our simulation's effectiveness at accurately modeling actual hockey 
games, we combined a dataset with the final event counts for 1500 actual hockey 
games and for 1500 simulated games. We then performed two experiments. First, We 
trained a KMeans, Logistic Regression, RandomForest and XGBoost model on this 
dataset with default parameters, to see 
if they could classify games as simulated or real. With a 70-30 train-test split, 
the accuracies on the test set were 47.9\%, 48.5\%, 67.5\% and 69.3\% respectively. 
Second, we fit and transformed this dataset using PCA, t-SNE and UMAP 
dimensionality reductions using various perplexity and neighbor hyperparameters to see 
if these algorithms clustered the synthetic games and actual games in 
different clusters. As shown in Figure~\ref{fig:simulation_v_actual}, the simulated
games and actual games are all clustered together. From these two experiments, we conclude
that our simulator effectively models live NHL games.

@@

\subsection{Method Comparison}
Each of our methods have one general purpose: predict the winner of the game regardless of the point of the game. 
To evaluate our models, we randomly selected 100 games outside of the training data. 
To ensure that the test set represented the actual data, we selected the test set with 50 home wins and 50 away wins, 
and we also made sure there was an even representation of each of the 32 NHL teams. 
We then used each model to predict a winner at various points throughout each game. 
From these predictions on the 100 games, we computed each model's accuracy for each point in the game.
